---
title: "Santa Barbara Breeding Bird Study: Graphs"
author: "Linus Blomqvist"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

```{r include = FALSE}
library(tidyverse)
library(lubridate)
library(gridExtra)
```

```{r}
# Read in data
data <- read.csv("BBS_data.csv")

# Change date for erroneous Allen's Hummingbird observation
data$Date[data$Record.Number == "1043"] <- as.Date("2006-02-10")

# Change data for Red-tailed Hawk observation
data$Date[data$Record.Number == "2441"] <- as.Date("2009-04-07")

# Delete erroneous record for RTHA
data <- data[data$Record.Number != 7170,]

# Standardize date format
data$Date <- dmy(data$Date)

# Make a new variables for week, month, day
data$Week <- week(data$Date)
data$Month <- month(data$Date)
data$Day <- day(data$Date)

# Get unique types of breeding evidence
breeding_evidence <- unique(data$Breeding.Evidence)

# Note: Index 5 and 27 both refer to fledgling under parental care

# Week and month key
week_key <- data.frame(floor_day = seq(as.Date("2021-1-1"), as.Date("2021-12-31"),
                                       by = "weeks"), Week = 1:53)
month_key <- data.frame(floor_day = seq(as.Date("2021-1-1"), as.Date("2021-12-31"),
                                        by = "months"), Month = 1:12)

week_key_Dec <- data.frame(floor_day = seq(as.Date("2020-12-1"), as.Date("2021-11-30"),
                                       by = "weeks"), Week = c(50:53, 1:49))
month_key_Dec <- data.frame(floor_day = seq(as.Date("2020-12-1"), as.Date("2021-11-30"),
                                        by = "months"), Month = c(12,1:11))

# Floor day is just a date in 2021 that corresponds to a given week (1-53)
# or month (1-12). This is so that the x axis retains a date format
```

```{r}
# Nojoqui Falls data on Starlings and Purple Martins
nojoqui <- rbind(read.csv("PUMA.csv"), read.csv("EUST.csv"))

# Make dummy in main dataset for Nojoqui records
data$Nojoqui <- ifelse(data$Record.Number %in% nojoqui$Record.Number, 1, 0)

# Find erroneous observation
puma_error <- data %>%
  filter(Common.Name == "Purple Martin") %>%
  filter(Month == 3) %>%
  select(Record.Number) %>%
  as.numeric()

# Delete this observation
data <- data[data$Record.Number != puma_error,]
```

```{r}
# Theme template
standard_theme <- theme(panel.grid.major.x = element_blank(),
          panel.grid.minor.x = element_blank(),
          text = element_text(size = 15),
          plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5))
```


```{r}
# Single species
single_species_plot_function <- function(species, breeding_evidence, time_aggr, time_key) {
  n_obs <- nrow(filter(data, Common.Name == species, Breeding.Evidence %in% breeding_evidence))
  data %>% # start with full dataset
  filter(Common.Name == species) %>% # only include select species
  filter(Breeding.Evidence %in% breeding_evidence) %>% # only certain kinds of evidence
  group_by(!!as.name(time_aggr)) %>% # summarize over weeks or months
  tally() %>% # count the number of observations
  left_join(time_key, by = time_aggr) %>% # merge a date onto the week/month number
  ggplot(aes(x = floor_day, y = n)) +
  geom_point() +
  labs(title = "Breeding Phenology",
       subtitle = species,
       x = "Date",
       y = str_c("Observations Per ", time_aggr),
       caption = str_c("n = ", n_obs, sep = "")) +
    theme_light() +
    standard_theme +
    theme(legend.position = "none") +
    scale_x_date(date_labels = "%b", date_breaks = "1 month")
}
```

Scaly-breasted Munia: all evidence used
```{r}
# Scaly-breasted Munia
single_species_plot_function("Scaly-breasted Munia", 
              breeding_evidence,
              "Month",
              month_key) +
  geom_smooth(span = 0.7, # choose smoothness of the line
              se = FALSE,
              color = "black")
```

Eurasia Collared-Dove: all evidence used
```{r}
# Eurasian Collared-Dove
single_species_plot_function("Eurasian Collared-Dove", 
              breeding_evidence,
              "Month",
              month_key) +
  scale_y_continuous(breaks = c(0, 3, 6, 9))
```

```{r}
# Single species with multiple categories of evidence: Red-tailed Hawk
# This is not yet generalizable (work in progress!)

# Select breeding evidence categories
rtha_evidence <- breeding_evidence[c(22,4,3,1,5,9,15,18,27)]

# Group this into two
data$rtha_evidence_cat <- rep(NA, nrow(data))
data$rtha_evidence_cat[data$Breeding.Evidence %in% breeding_evidence[c(22,4,3)]] <- "Pre-fledge"
data$rtha_evidence_cat[data$Breeding.Evidence %in% breeding_evidence[c(1,5,9,15,18,27)]] <- "Post-fledge"
data$rtha_evidence_cat <- factor(data$rtha_evidence_cat, levels = c("Pre-fledge", "Post-fledge"))

# Plot function
single_species_mult_ev_plot_function <- function(species, breeding_evidence, time_aggr, time_key) {
  data %>% # start with full dataset
  filter(Common.Name == species) %>% # only include select species
  filter(Breeding.Evidence %in% breeding_evidence) %>% # only certain kinds of evidence
  group_by(!!as.name(time_aggr), rtha_evidence_cat) %>% # summarize over weeks or months
  tally() %>% # count the number of observations
  left_join(time_key, by = time_aggr) %>% # merge a date onto the week/month number
  ggplot(aes(x = floor_day, y = n, color = rtha_evidence_cat)) +
  geom_point() +
  labs(title = "Breeding Phenology",
       subtitle = species,
       x = "Date",
       y = str_c("Observations Per ", time_aggr),
       caption = str_c("n = ", nrow(filter(data, Common.Name == species, 
                          Breeding.Evidence %in% breeding_evidence)), sep = ""),
       color = element_blank()) +
    theme_light() +
    standard_theme +
    theme(legend.position = "bottom") +
    scale_x_date(date_labels = "%b", date_breaks = "1 month") +
    scale_color_brewer(palette = "Dark2")
}
```

Red-tailed Hawk evidence used:

```{r}
rtha_evidence
```

```{r}
# Plot
single_species_mult_ev_plot_function("Red-tailed Hawk",
                                     rtha_evidence,
                                     "Week",
                                     week_key) +
    geom_smooth(span = 1, # choose smoothness of the line
              se = FALSE)
```


```{r}
# Pair with different evidence
# Species pair
species_pair_plot_function <- function(dataset, species_1, species_2, breeding_evidence_1,
                                       breeding_evidence_2, time_aggr, time_key) {
  df_species_1 <- dataset %>% # select data for species 1
  filter(Common.Name == species_1) %>%
  filter(Breeding.Evidence %in% breeding_evidence_1)

df_species_2 <- dataset %>% # select data for species 2
  filter(Common.Name == species_2) %>%
  filter(Breeding.Evidence %in% breeding_evidence_2)

df_both <- rbind(df_species_1, df_species_2) # merge data for both species

# Create caption with number of observations for each species
n_obs_1 <- nrow(filter(data, Common.Name == species_1, Breeding.Evidence %in% breeding_evidence_1))
n_obs_2 <- nrow(filter(data, Common.Name == species_2, Breeding.Evidence %in% breeding_evidence_2))
n_caption <- str_c(species_1, " n = ", n_obs_1, "; ", species_2, " n = ", n_obs_2)

  df_both %>%
  group_by(!!as.name(time_aggr), Common.Name) %>% # summarize over weeks or months, and by species name
  tally() %>% # count the number of observations
  left_join(time_key, by = time_aggr) %>% # merge a date onto the week/month number
  ggplot(aes(x = floor_day, y = n, color = Common.Name)) +
  geom_point() +
  labs(title = "Breeding Phenology",
       x = "Date",
       y = str_c("Observations Per ", time_aggr),
       color = element_blank(),
       caption = n_caption,
       subtitle = str_c(species_1, " and ", species_2)) +
    theme_light() +
    standard_theme +
    theme(legend.position = "bottom") +
    scale_x_date(date_labels = "%b", date_breaks = "1 month") +
    scale_color_brewer(palette = "Set1")
}
```


```{r}
# Mallard vs Gadwall
evidence_mallard <- breeding_evidence[c(16,4,7,9,1,5,18,20,6,3,27)]
evidence_gadwall <- breeding_evidence[c(16,4,7,5,18,3,27)]

species_pair_plot_function(data,
                           "Gadwall", 
                           "Mallard", 
                           evidence_mallard,
                           evidence_gadwall,
                           "Week", 
                           week_key) +
    geom_smooth(span = 1, # choose smoothness of the line
              se = FALSE)
```

Mallard evidence used:

```{r}
evidence_mallard
```

Gadwall evidence used:

```{r}
evidence_gadwall
```


```{r}
# Red-tailed vs Red-shouldered Hawk
evidence_rtha <- breeding_evidence[c(16,13,10,7,9,15,1,5,18,12,3,21,17,27)]
evidence_rsha <- breeding_evidence[c(16,13,10,9,15,1,5,18,12,3,17,27)]

species_pair_plot_function(data,
                           "Red-shouldered Hawk", 
                           "Red-tailed Hawk", 
                           evidence_rtha,
                           evidence_rsha, 
                           "Week", 
                           week_key_Dec) +
    geom_smooth(span = 0.7, # choose smoothness of the line
              se = FALSE)
```

Red-shouldered Hawk evidence used:

```{r}
evidence_rsha
```

Red-tailed Hawk evidence used:

```{r}
evidence_rtha
```





```{r}
# Anna's vs Allen's Hummingbird
evidence_anhu <- breeding_evidence[c(16,13,22,10,7,9,15,1,5,18,12,3,21,17,27)]
evidence_alhu <- breeding_evidence[c(16,13,22,10,9,15,1,5,18,12,3,17,27)]

species_pair_plot_function(data,
                           "Allen's Hummingbird", 
                           "Anna's Hummingbird", 
                           evidence_anhu,
                           evidence_alhu, 
                           "Week", 
                           week_key_Dec) +
    geom_smooth(span = 0.7, # choose smoothness of the line
              se = FALSE)
```

Allen's Hummingbird evidence used:

```{r}
evidence_alhu
```

Anna's hummingbird evidence used:

```{r}
evidence_anhu
```



```{r}
# Starling and Purple Martin at Nojoqui

# Need to extract the types of breeding evidence for both
eust_evidence <- unique(nojoqui$Breeding.Evidence[nojoqui$Common.Name == "European Starling"])
puma_evidence <- unique(nojoqui$Breeding.Evidence[nojoqui$Common.Name == "Purple Martin"])

species_pair_plot_function(filter(data, Nojoqui == 1),
                           "European Starling", 
                           "Purple Martin", 
                           eust_evidence,
                           puma_evidence, 
                           "Week", 
                           week_key_Dec) +
    geom_smooth(span = 0.8, # choose smoothness of the line
              se = FALSE)

# The code currently generates the n for Starling that corresponds to all records in the county.
# I need to include an option for Nojoqui in the function for a future iteration
```

European Starling evidence used:

```{r}
eust_evidence
```

Purple Martin evidence used:

```{r}
puma_evidence
```



```{r}
# Group of birds eg native species, raptors

species_group_plot_function <- function(species, breeding_evidence, time_aggr, time_key) {
  n_obs <- nrow(filter(data, Common.Name %in% species, Breeding.Evidence %in% breeding_evidence))
  data %>% # start with full dataset
  filter(Common.Name %in% species) %>% # only include select species
  filter(Breeding.Evidence %in% breeding_evidence) %>% # only certain kinds of evidence
  group_by(!!as.name(time_aggr)) %>% # summarize over weeks or months
  tally() %>% # count the number of observations
  left_join(time_key, by = time_aggr) %>% # merge a date onto the week/month number
  ggplot(aes(x = floor_day, y = n)) +
  geom_point(color = "black") +
  labs(title = "Breeding Phenology",
       x = "Date",
       y = str_c("Observations Per ", time_aggr),
       caption = str_c("n = ", n_obs, sep = "")) +
    theme_light() +
    standard_theme +
    theme(legend.position = "none") +
    scale_x_date(date_labels = "%b", date_breaks = "1 month")
}
```

```{r}
# All native
all_species <- unique(data$Common.Name)
native_species <- all_species[!(all_species %in% c("European Starling", "Swan Goose",
                                                 "Eurasian Collared-Dove"))]

species_group_plot_function(native_species,
                            breeding_evidence,
                            "Week", # could also choose month
                            week_key) +
  geom_smooth(span = 0.3, # choose smoothness of the line
              se = FALSE,
              color = "black",
              lwd = 0.8) + # if month above, put in "month_key"
  labs(subtitle = "All native species")
```

All native species evidence used: all

```{r}
# Raptors
raptors <- c("American Kestrel", "Cooper's Hawk", "Northern Goshawk", 
             "Northern Harrier", "Peregrine Falcon", "Red-shouldered Hawk",
             "Red-tailed Hawk", "Sharp-shinned Hawk", "Swainson's Hawk",
             "Turkey Vulture", "White-tailed Kite")

raptor_evidence <- breeding_evidence[c(16,13,14,23,10,4,7,15,1,5,18,6,12,3,27)]

species_group_plot_function(raptors,
                            raptor_evidence,
                            "Week", # could also choose month
                            week_key_Dec) +
  geom_smooth(span = 0.6, # choose smoothness of the line
              se = FALSE,
              color = "black",
              lwd = 0.8) + # if month above, put in "month_key"
  labs(subtitle = "All raptors except owls and eagles")
```

All raptors evidence used:

```{r}
raptor_evidence
```


```{r}
# Violin plot with all native species
data$breeding_intensity <- 1
data %>%
  left_join(week_key, by = "Week") %>%
  ggplot(aes(x = breeding_intensity, y = floor_day)) + 
  geom_violin(fill = "lightblue") +
  coord_flip() +
  theme_light() +
  standard_theme +
  theme(axis.text.y=element_blank()) +
  labs(title = "Breeding Season",
       x = "Relative breeding intensity",
       y = "Date",
       caption = "A wider plot indicates a larger number of breeding observations in a given week") +
  scale_y_date(date_labels = "%b", date_breaks = "1 month")
```
