---
title: "Santa Barbara Breeding Bird Study"
author: "Linus Blomqvist"
date: "3/2/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r include = FALSE}
library(tidyverse)
library(lubridate)
library(gridExtra)
```

## Introduction
Here, I explore ways to present data from the Santa Barbara Breeding Bird Study. Since this is exploratory, it should not be cited.

```{r}
# Read in data
data_full <- read.csv("BBS_data.csv")
```

There are `r length(unique(data_full$Breeding.Evidence))` types of breeding evidence in the dataset. We are focusing on 15 of them in the following five categories:

* <span style="color:red">Nest construction</span>
  + <span style="color:red">Carrying nest material</span>
  + <span style="color:red">Nest building</span>
* <span style="color:blue">Copulation and eggs in nest</span>
  + <span style="color:blue">Copulation</span>
  + * <span style="color:blue">Egg in nest</span>
* <span style="color:black">Nestling and brancher</span>
  + <span style="color:black">Carrying food</span>
  + <span style="color:black">Delivering food to nest or cavity</span>
  + <span style="color:black">Nestling in nest</span>
  + <span style="color:black">Carrying fecal sac</span>
  + <span style="color:black">Fledgling out of nest -- brancher</span>
* <span style="color:green">Fledglings and families</span>
  + <span style="color:green">Family group in close association</span>
  + <span style="color:green">Feeding fledgling</span>
  + <span style="color:green">Feedling begging</span>
  + <span style="color:green">Fledgling under parental care</span>
  + <span style="color:green">Fledgling with presumed parent</span>
* <span style="color:orange">Juvenile independent</span>

## Data prep
In the code below, I group the evidence types in the dataset into those categories, labeled by the corresponding color in the previous list, and then make a categorical variable for the five groups.


```{r}
# Get unique types of breeding evidence
breeding_evidence_cat <- unique(data_full$Breeding.Evidence)

# Select the 15 Mark is interested in
select_evidence_cat <- breeding_evidence_cat[c(11, 6, 22, 4, 13, 10, 3, 24, 1, 7, 9, 15, 27, 18, 20)]

# Group into the five categories
red_cat <- select_evidence_cat[c(1,2)]
blue_cat <- select_evidence_cat[c(3,4)]
black_cat <- select_evidence_cat[5:9]
green_cat <- select_evidence_cat[10:14]
orange_cat <- select_evidence_cat[15]

# Filter the dataset to only include observations with evidence in one of the 15 categories
data <- data_full %>% filter(Breeding.Evidence %in% select_evidence_cat)

# Make categorical variable
data$evidence_category <- rep(NA, nrow(data))
data$evidence_category[data$Breeding.Evidence %in% red_cat] <- "Nest construction"
data$evidence_category[data$Breeding.Evidence %in% blue_cat] <- "Copulation and eggs in nest"
data$evidence_category[data$Breeding.Evidence %in% black_cat] <- "Nestling and brancher"
data$evidence_category[data$Breeding.Evidence %in% green_cat] <- "Fledglings and families"
data$evidence_category[data$Breeding.Evidence %in% orange_cat] <- "Juvenile independent"

# Standardize date format
data$Date <- dmy(data$Date)

# Make a new variable, week
data$week <- week(data$Date)
data$year <- year(data$Date)
data$month <- month(data$Date)
data$day <- day(data$Date)

# Group by and summarize by week
breeding_evidence_by_week <- data %>%
  group_by(week, year, Common.Name, evidence_category) %>%
  tally()

# Group by and summarize by day
breeding_evidence_by_day <- data %>%
  group_by(day, month, Common.Name, evidence_category) %>%
  tally()

breeding_evidence_by_week$evidence_category <- factor(breeding_evidence_by_week$evidence_category, levels = c("Nest construction", "Copulation and eggs in nest", "Nestling and brancher", "Fledglings and families", "Juvenile independent"))

breeding_evidence_by_day$evidence_category <- factor(breeding_evidence_by_day$evidence_category, levels = c("Nest construction", "Copulation and eggs in nest", "Nestling and brancher", "Fledglings and families", "Juvenile independent"))


# Concatenate to date of year
breeding_evidence_by_day$ymd <- str_c("2021", breeding_evidence_by_day$month, 
                                      breeding_evidence_by_day$day, sep = "-")
breeding_evidence_by_day$date <- ymd(breeding_evidence_by_day$ymd)
```

## Figure templates

```{r}
# Density
density_plot <- function(species) {
  ggplot(filter(breeding_evidence_by_day, Common.Name == species), aes(x = date, fill = evidence_category)) +
    geom_density(adjust=1.5, alpha=.3) +
    labs(title = "Breeding Phenology", 
       subtitle = species) +
    labs(x = "Date",
         y = "Density",
         fill = "Breeding evidence") +
    theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        # panel.background = element_rect(fill='white', colour='black'),
        # panel.grid.major.y = element_line(colour = "gray"),
        legend.position = "right",
        axis.text.y = element_blank(),
        panel.background = element_blank(),
        axis.ticks.y = element_blank()) +
    scale_x_date(date_labels = "%b", date_breaks = "1 month")
}

# Density for comparison between two species with facet WORK IN PROGRESS
density_comparison <- function(species_list, evidence_categories) {
  breeding_evidence_by_day %>%
    filter(Common.Name %in% species_list,
         evidence_category %in% evidence_categories) %>% # among select_evidence_cat
    ggplot(aes(x = date, fill = evidence_category)) +
    geom_density(adjust=1.5, alpha=.3) +
    facet_grid(rows = vars(Common.Name)) +
    labs(title = "Breeding Phenology") +
    labs(x = "Date",
         y = "Density",
         fill = "Breeding evidence") +
    theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        # panel.background = element_rect(fill='white', colour='black'),
        # panel.grid.major.y = element_line(colour = "gray"),
        legend.position = "right",
        axis.text.y = element_blank(),
        panel.background = element_blank(),
        axis.ticks.y = element_blank()) +
    scale_x_date(date_labels = "%b", date_breaks = "1 month")
}



# Smoothed means
smoothed_means_plot <- function(species) {
  ggplot(filter(breeding_evidence_by_day, Common.Name == species), 
         aes(x = date, y = n, color = evidence_category)) +
  geom_point() +
  geom_smooth(se = FALSE, 
              lwd = 0.8) +
  scale_color_manual(values = c("red", "blue", "black", "green", "orange")) +
  labs(x = "Date", 
       y = "Count", 
       color = "Breeding evidence") +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        panel.background = element_rect(fill='white', colour='black'),
        panel.grid.major.y = element_line(colour = "gray"),
        legend.position = "right") +
  labs(title = "Breeding Phenology", 
       subtitle = species, 
       caption = "Dots represent total observations in a given week. Lines represent smoothed means.") +
    scale_x_date(date_labels = "%b", date_breaks = "1 month")
}

# Boxplot
boxplot_bird <- function(species) {
  ggplot(filter(breeding_evidence_by_day, Common.Name == species), aes(x = evidence_category, y = date)) +
  geom_boxplot() +
  geom_jitter(color = "blue", width = 0.05, size = 0.3, alpha = 0.9) +
  coord_flip() +
  labs(title = "Breeding Phenology", 
       subtitle = species) +
  labs(x = "Breeding evidence", 
       y = "Date",
       caption = "Dots indicate observations") +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        panel.background = element_rect(fill='white', colour='black'),
        panel.grid.major.y = element_line(colour = "gray"),
        legend.position = "right")
}
```

```{r}
density_plot("Red-tailed Hawk")
```

```{r}
density_comparison("Mallard", "Gadwall")
```



# Misc
```{r eval = FALSE}
jpeg("figures/rplot.jpg")
# 2. Create the plot
density_plot
# 3. Close the file
dev.off()
```

```{r eval = FALSE}

# loop to produce figures for all species - not complete
species_names <- unique(data$Common.Name)

for (i in 1:length(species_names)) {
  
}
ggplot(filter(breeding_evidence, Common.Name == species_names[i]), aes(x = week, color = evidence_category, fill = evidence_category)) +
    geom_density(adjust=1.5, alpha=.3)
```

